A generalization of the Bradley-Terry model for draws in chess

### 2020-09-14 inst/jebo_chess_rebuttal.txt

"A generalization of the Bradley-Terry model for draws in chess"

JEBO-D-20-00916

Referee Report

This paper develops extensions to the Bradley-Terry model appropriate
to Bernoulli trials so that the model can accommodate the
chess-specific needs of draws and the first-mover advantage possessed
by White. The author takes this extended model to various sets of
exclusively game-level outcomes and estimates time-invariant player
strengths as well as the importance of draws and first-mover
advantage. After estimating common parameters for draws and
first-mover advantage, the author allows for (time-invariant)
player-specific parameters on these margins. The paper then concludes
by considering distinct draw parameters for Soviet and non-Soviet
players. Using data from the top chess events of 1962 and 1963, the
author determines that Soviet players draw more frequently against
each other than in other game scenarios.

The paper does a nice job of clearly laying out the various reified
"monsters" that extend the original Bradley-Terry model to accommodate
chess. This framework is plausible and intuitive. I also appreciated
the paper's brevity in making its point. The allowance for
playerspecific tendencies to draw or to gain excess-advantage from
moving first also strikes me as an innovation to the
literature. Lastly, the paper is well-written and a fun read.

My comments on the rest of the paper are detailed below by decreasing
importance.

1) The author never properly explores the paper's identification
strategy. As I read it, the identifying assumption, besides the
functional form assumption, is that player strengths are constant over
the duration of the considered data. This seems highly plausible
within a specific event (e.g., the Candidates tournament of 1963, the
Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
example. The author does not specify the timespan over which these 365
games were played, but I imagine that the timespan must exceed 10
years. To the extent that all players have matured to full strength
and none have yet begun to decline, this assumption would be benign,
but it should be supported.

Precise estimation will hinge on a large number of games, but this
will be in tension with the assumption of time-invariant
player-strengths. The author should explore this tension.

2) The author neglects to motivate the paper beyond its application
within the chess world.  The tournament literature's applications lie
primarily, though definitely not exclusively, within compensation and
labor. How could this extension be used within that context?  Where
might economists see the first-mover advantage in action? Where might
the extension to allow draws be especially lucrative? Etc.

3) As the author notes, the work of Moul and Nye (2009) is directly on
point. The author's discussion of this paper, however, is overly
sparse and at times misleading. Moul and Nye use existing observed
player strengths (variations of commonly used scores based on the Elo
rating system that goes back to 1960). Moul and Nye then use these
strengths within an ordered probit to estimate parameters that relate
back to draw likelihood and White's first-mover advantage. The sharp
difference between that paper and this is then that the author is
estimating player relative strengths (conditional upon a pool of
players) while Moul and Nye use "observed" strengths [footnote: Those
relative strengths are highly predictive, with t-statistics of around
20]. The latter approach is far less parametrically demanding, but
leans much more on the assumption of normally distributed disturbances
(basically the differing assumption of functional form). The latter
approach also allows individuals' player strength to vary across time
and is far more flexible to changes in the composition of the pool of
players.

Furthermore, Moul and Nye estimate several versions of their ordered
probit without data on the number of moves (ply-count). This is
contrary to the implication of the paper (p. 6, immediately below
Table 2), which suggests that the ply-count data are essential to that
paper's findings. All of that paper's core results (significance of
draws, first-mover advantage, a distinct draw outcome for round-robin
international tournament games that involve two Soviets) are
originally derived without ply-count data. The ply-count data are
incorporated to explore a potential mechanism by which players in a
draw-cartel might benefit.

4) The paper concludes that the estimates derived from the 1962
Interzonals and the 1963 Candidates strongly suggest that the "draw
monster" won games involving two Soviet players more frequently than
games that involved at least one non-Soviet. This is consistent with a
Soviet cartel, but it could also be the result of Soviet-specific
styles of play. Moul and Nye (2009) test the Soviet style-of-play
hypothesis by considering international tournaments when the format
changed from all-play-all to knockout and allSoviet tournaments. Both
comparisons yield results consistent with a Soviet draw-cartel in
international all-play-all tournaments and rejecting the Soviet
style-of-play hypothesis.  By only considering tournaments that were
international and all-play-all, the author forgoes this critical
source of identification.

In the cartel example, the author also neglects the advantage that
presumably accrued to Soviet players when they played non-Soviet
players in all-play-all international tournaments. This would suggest
another "monster" to be estimated, one that benefits the Soviet player
when playing a non-Soviet opponent. Without such an addition, Soviet
strengths will be overstated.

### 2020-09-14 inst/jebo_chess_rebuttal.txt

3) As the author notes, the work of Moul and Nye (2009) is directly on
point.  The author's discussion of this paper, however, is overly
sparse and at times misleading.  Moul and Nye use existing observed
player strengths (variations of commonly used scores based on the Elo
rating system that goes back to 1960).  Moul and Nye then use these
strengths within an ordered probit to estimate parameters that relate
back to draw likelihood and White's first-mover advantage.  The sharp
difference between that paper and this is then that the author is
estimating player relative strengths (conditional upon a pool of
players) while Moul and Nye use "observed" strengths [footnote: Those
relative strengths are highly predictive, with t-statistics of around
20].  The latter approach is far less parametrically demanding, but
leans much more on the assumption of normally distributed disturbances
(basically the differing assumption of functional form).  The latter
approach also allows individuals' player strength to vary across time
and is far more flexible to changes in the composition of the pool of
players.

Furthermore, Moul and Nye estimate several versions of their ordered
probit without data on the number of moves (ply-count).  This is
contrary to the implication of the paper (p. 6, immediately below
Table 2), which suggests that the ply-count data are essential to that
paper's findings.  All of that paper's core results (significance of
draws, first-mover advantage, a distinct draw outcome for round-robin
international tournament games that involve two Soviets) are
originally derived without ply-count data.  The ply-count data are
incorporated to explore a potential mechanism by which players in a
draw-cartel might benefit.

4) The paper concludes that the estimates derived from the 1962
Interzonals and the 1963 Candidates strongly suggest that the "draw
monster" won games involving two Soviet players more frequently than
games that involved at least one non-Soviet.  This is consistent with
a Soviet cartel, but it could also be the result of Soviet-specific
styles of play.  Moul and Nye (2009) test the Soviet style-of-play
hypothesis by considering international tournaments when the format
changed from all-play-all to knockout and all Soviet tournaments.
Both comparisons yield results consistent with a Soviet draw-cartel in
international all-play-all tournaments and rejecting the Soviet
style-of-play hypothesis.  By only considering tournaments that were
international and all-play-all, the author forgoes this critical
source of identification.

### 2020-09-14 inst/jebo_chess_rebuttal.txt

"A generalization of the Bradley-Terry model for draws in chess"

JEBO-D-20-00916

Referee Report

> This paper develops extensions to the Bradley-Terry model appropriate
> to Bernoulli trials so that the model can accommodate the
> chess-specific needs of draws and the first-mover advantage possessed
> by White.  The author takes this extended model to various sets of
> exclusively game-level outcomes and estimates time-invariant player
> strengths as well as the importance of draws and first-mover
> advantage.  After estimating common parameters for draws and
> first-mover advantage, the author allows for (time-invariant)
> player-specific parameters on these margins.  The paper then concludes
> by considering distinct draw parameters for Soviet and non-Soviet
> players.  Using data from the top chess events of 1962 and 1963, the
> author determines that Soviet players draw more frequently against
> each other than in other game scenarios.
> 
> The paper does a nice job of clearly laying out the various reified
> "monsters" that extend the original Bradley-Terry model to accommodate
> chess.  This framework is plausible and intuitive.  I also appreciated
> the paper's brevity in making its point. The allowance for
> playerspecific tendencies to draw or to gain excess-advantage from
> moving first also strikes me as an innovation to the literature.
> Lastly, the paper is well-written and a fun read.
> 
> 
> My comments on the rest of the paper are detailed below by decreasing
> importance.
> 
> 1) The author never properly explores the paper's identification
> strategy.  As I read it, the identifying assumption, besides the
> functional form assumption, is that player strengths are constant over
> the duration of the considered data. This seems highly plausible
> within a specific event (e.g., the Candidates tournament of 1963, the
> Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
> example.  The author does not specify the timespan over which these
> 365 games were played, but I imagine that the timespan must exceed 10
> years.  To the extent that all players have matured to full strength
> and none have yet begun to decline, this assumption would be benign,
> but it should be supported.
> 
> Precise estimation will hinge on a large number of games, but this
> will be in tension with the assumption of time-invariant
> player-strengths.  The author should explore this tension.
> 
> 2) The author neglects to motivate the paper beyond its application
> within the chess world.  The tournament literature's applications lie
> primarily, though definitely not exclusively, within compensation and
> labor.  How could this extension be used within that context?  Where
> might economists see the first-mover advantage in action? Where might
> the extension to allow draws be especially lucrative? Etc.
> 
> 3) As the author notes, the work of Moul and Nye (2009) is directly on
> point.  The author's discussion of this paper, however, is overly
> sparse and at times misleading.  Moul and Nye use existing observed
> player strengths (variations of commonly used scores based on the Elo
> rating system that goes back to 1960).  Moul and Nye then use these
> strengths within an ordered probit to estimate parameters that relate
> back to draw likelihood and White's first-mover advantage.  The sharp
> difference between that paper and this is then that the author is
> estimating player relative strengths (conditional upon a pool of
> players) while Moul and Nye use "observed" strengths [footnote: Those
> relative strengths are highly predictive, with t-statistics of around
> 20].  The latter approach is far less parametrically demanding, but
> leans much more on the assumption of normally distributed disturbances
> (basically the differing assumption of functional form).  The latter
> approach also allows individuals' player strength to vary across time
> and is far more flexible to changes in the composition of the pool of
> players.
> 
> Furthermore, Moul and Nye estimate several versions of their ordered
> probit without data on the number of moves (ply-count).  This is
> contrary to the implication of the paper (p. 6, immediately below
> Table 2), which suggests that the ply-count data are essential to that
> paper's findings.  All of that paper's core results (significance of
> draws, first-mover advantage, a distinct draw outcome for round-robin
> international tournament games that involve two Soviets) are
> originally derived without ply-count data.  The ply-count data are
> incorporated to explore a potential mechanism by which players in a
> draw-cartel might benefit.
> 
> 4) The paper concludes that the estimates derived from the 1962
> Interzonals and the 1963 Candidates strongly suggest that the "draw
> monster" won games involving two Soviet players more frequently than
> games that involved at least one non-Soviet.  This is consistent with
> a Soviet cartel, but it could also be the result of Soviet-specific
> styles of play.  Moul and Nye (2009) test the Soviet style-of-play
> hypothesis by considering international tournaments when the format
> changed from all-play-all to knockout and all Soviet tournaments.
> Both comparisons yield results consistent with a Soviet draw-cartel in
> international all-play-all tournaments and rejecting the Soviet
> style-of-play hypothesis.  By only considering tournaments that were
> international and all-play-all, the author forgoes this critical
> source of identification.
> 
> In the cartel example, the author also neglects the advantage that
> presumably accrued to Soviet players when they played non-Soviet
> players in all-play-all international tournaments.  This would suggest
> another "monster" to be estimated, one that benefits the Soviet player
> when playing a non-Soviet opponent.  Without such an addition, Soviet
> strengths will be overstated.
>

### 2020-09-14 inst/jebo_chess_rebuttal.txt

"A generalization of the Bradley-Terry model for draws in chess"

JEBO-D-20-00916

Referee Report

The referee's comments are preceded by a chevron (">") and my response
is included inline.  I have changed line-breaks in the interests of
clarity.

Overall Response

I am delighted that this insightful referee "gets"
the manuscript.  The referee makes a number of observations and
suggestions, all of which I agree with, and I have made changes to the
submission in response.  I have tried very hard to preserve the
brevity and pleasing structure that the referee explicitly notes.  In
short, the revised manuscript is a stronger and more widely applicable
document and I commend it to you.

Software implementation note

In the interests of full disclosure, I have made a significant change
to the C++ back end of the hyper2 package:

https://github.com/RobinHankin/hyper2/commit/d9d16c45daa752d4687446d7df9805439f461451#diff-25d902c24283ab8cfbac54dfa101ad31

The hyper2 package now uses player names as keys in the STL map class,
rather than integers.  The change is a conceptual streamlining which
results in simplified R idiom for a number of cases, but does not
change the mathematical basis of the software or alter any numerical
results.

Detailed responses

### 2020-09-15 inst/jebo_chess_rebuttal.txt

Referee Report

The referee's comments are preceded by a chevron (">") and my response
is included inline.  I have changed line-breaks in the interests of
clarity.

Overall Response

I am delighted that this insightful referee "gets" the manuscript.
The referee makes a number of observations and suggestions, all of
which I agree with, and I have made changes to the submission in
response, marked in red.  I have tried very hard to preserve the
brevity and pleasing structure that the referee explicitly notes.  In
short, the revised manuscript is a stronger and more widely applicable
document and I commend it to you.

@@

https://github.com/RobinHankin/hyper2/commit/d9d16c45daa752d4687446d7df9805439f461451#diff-25d902c24283ab8cfbac54dfa101ad31

The hyper2 package now uses player names as keys in the STL map class,
rather than integers.  The change is a conceptual streamlining which
results in simplified R idiom for a number of cases, but does not
change the mathematical basis of the software or alter any numerical
results.

Detailed responses

> This paper develops extensions to the Bradley-Terry model appropriate
> to Bernoulli trials so that the model can accommodate the
> chess-specific needs of draws and the first-mover advantage possessed
> by White.  The author takes this extended model to various sets of
> exclusively game-level outcomes and estimates time-invariant player
> strengths as well as the importance of draws and first-mover
> advantage.  After estimating common parameters for draws and
> first-mover advantage, the author allows for (time-invariant)
> player-specific parameters on these margins.  The paper then concludes
> by considering distinct draw parameters for Soviet and non-Soviet
> players.  Using data from the top chess events of 1962 and 1963, the
> author determines that Soviet players draw more frequently against
> each other than in other game scenarios.
> 
> The paper does a nice job of clearly laying out the various reified
> "monsters" that extend the original Bradley-Terry model to accommodate
> chess.  This framework is plausible and intuitive.  I also appreciated
> the paper's brevity in making its point. The allowance for
> playerspecific tendencies to draw or to gain excess-advantage from
> moving first also strikes me as an innovation to the literature.
> Lastly, the paper is well-written and a fun read.
> 
> 
> My comments on the rest of the paper are detailed below by decreasing
> importance.
> 
> 1) The author never properly explores the paper's identification
> strategy.  As I read it, the identifying assumption, besides the
> functional form assumption, is that player strengths are constant over
> the duration of the considered data. This seems highly plausible
> within a specific event (e.g., the Candidates tournament of 1963, the
> Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
> example.  The author does not specify the timespan over which these
> 365 games were played, but I imagine that the timespan must exceed 10
> years.  To the extent that all players have matured to full strength
> and none have yet begun to decline, this assumption would be benign,
> but it should be supported.

An excellent observation.  In the chess world the pre-eminent strength
measure would be the Elo rating in which the winning player takes
points from the loser based on the pre-match difference in Elo rating.
Elo exhibits an emergent phenomenon in which sudden changes of ability
are not reflected instantly but have an exponentially damped response.

My view is that this damping feature is sometimes desirable (old games
are surely less informative than more recent ones) but uncontrolled in
the sense that the time-constant is not explicit or controllable, and
the Elo system uses an arbitrary points system with arbitrary cut-offs
[as opposed to the objective Bradley-Terry system].

> 
> Precise estimation will hinge on a large number of games, but this
> will be in tension with the assumption of time-invariant
> player-strengths.  The author should explore this tension.
>

I am in total agreement with this observation.  My thinking on this
issue has been strongly affected by my recent work applying
Bradley-Terry to Formula 1 motor racing.  In F1, drivers generally
start their careers as weak competitors, become stronger with
experience, and finally become uncompetitive due to age [although
frankly Lewis Hamilton at 35yo seems to buck this general trend, as
recent events in Tuscany would indicate].  I would suggest that chess
players have a similar career arc, although Formula 1 Teams typically
take a dim view of underperforming drivers---whom they rapidly and
unceremonially drop.  The career arc for these competitors would have
a truncated right hand side, unlike chess players who typically
decline gradually over decades [although I would observe that
Kasparov's current FIDE rating of 2812 is not that much less than his
peak of 2851].

Specifically, the dataset in question stretches from 1981 to 2017, a
very wide range.  Generalized Bradley-Terry, together with reified
entities, is capable of handling strengths varying over a career: one
would introduce a "mid-career monster", or perhaps an "experience
monster".  I would add here that monsters' strength is inherently
positive, so one could deal with waning strength with age by
introducing a "youth monster" who helps younger players.  But such
analysis is, I would suggest, firmly in the category of further work.
I have added a discussion of this issue to the manuscript.

> 2) The author neglects to motivate the paper beyond its application
> within the chess world.  The tournament literature's applications lie
> primarily, though definitely not exclusively, within compensation and
> labor.  How could this extension be used within that context?  Where
> might economists see the first-mover advantage in action? Where might
> the extension to allow draws be especially lucrative? Etc.
> 
> 3) As the author notes, the work of Moul and Nye (2009) is directly on
> point.  The author's discussion of this paper, however, is overly
> sparse and at times misleading.  Moul and Nye use existing observed
> player strengths (variations of commonly used scores based on the Elo
> rating system that goes back to 1960).  Moul and Nye then use these
> strengths within an ordered probit to estimate parameters that relate
> back to draw likelihood and White's first-mover advantage.  The sharp
> difference between that paper and this is then that the author is
> estimating player relative strengths (conditional upon a pool of
> players) while Moul and Nye use "observed" strengths [footnote: Those
> relative strengths are highly predictive, with t-statistics of around
> 20].  The latter approach is far less parametrically demanding, but
> leans much more on the assumption of normally distributed disturbances
> (basically the differing assumption of functional form).  The latter
> approach also allows individuals' player strength to vary across time
> and is far more flexible to changes in the composition of the pool of
> players.
> 
> Furthermore, Moul and Nye estimate several versions of their ordered
> probit without data on the number of moves (ply-count).  This is
> contrary to the implication of the paper (p. 6, immediately below
> Table 2), which suggests that the ply-count data are essential to that
> paper's findings.  All of that paper's core results (significance of
> draws, first-mover advantage, a distinct draw outcome for round-robin
> international tournament games that involve two Soviets) are
> originally derived without ply-count data.  The ply-count data are
> incorporated to explore a potential mechanism by which players in a
> draw-cartel might benefit.
> 
> 4) The paper concludes that the estimates derived from the 1962
> Interzonals and the 1963 Candidates strongly suggest that the "draw
> monster" won games involving two Soviet players more frequently than
> games that involved at least one non-Soviet.  This is consistent with
> a Soviet cartel, but it could also be the result of Soviet-specific
> styles of play.  Moul and Nye (2009) test the Soviet style-of-play
> hypothesis by considering international tournaments when the format
> changed from all-play-all to knockout and all Soviet tournaments.
> Both comparisons yield results consistent with a Soviet draw-cartel in
> international all-play-all tournaments and rejecting the Soviet
> style-of-play hypothesis.  By only considering tournaments that were
> international and all-play-all, the author forgoes this critical
> source of identification.
> 
> In the cartel example, the author also neglects the advantage that
> presumably accrued to Soviet players when they played non-Soviet
> players in all-play-all international tournaments.  This would suggest
> another "monster" to be estimated, one that benefits the Soviet player
> when playing a non-Soviet opponent.  Without such an addition, Soviet
> strengths will be overstated.
>

### 2020-09-15 inst/jebo_chess_revision2.tex
begin addressing point 1

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

@@

>>>>

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}, \textcolor{red}{showing games played from 1981 to
  2017.  During this time, the players' strengths might have changed
  substantially.  Statistical assessment of time-varying strength is
  possible with the reified entity technique and this is discussed
  under further work below.  However, we may adopt one of two
  approaches: either we make the working assumption that the strengths
  do not change, or alternatively we wish to make inferences about the
  players' strength {\em averaged over a playing career}.  Either
  way,} Equation~\ref{likelihood_WD} becomes

@@

\textcolor{red}{One strong assumption made in the
  Kasparov-Karpov-Anand analysis was the constancy of the players'
  strengths over a 36-year period.  Many sports have a career arc
  whereby competitors start their careers weak, become stronger with
  experience, and finally become uncompetitive due to age.  Further
  work might include the introduction of additional reified entities
  that account for such phenomena: we might have a ``mid career
  monster'' who helps competitors at or near their career peak.}
